---
title: "Code"
author: "Clara Livingston & Emily Kaegi"
date: "May 25, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(lme4)
```

#Data Join
```{r}
state <- read.csv("StateData.csv")
team <- read.csv("TeamData2.csv",stringsAsFactors = TRUE)

statestuff <- data.frame(state.abb,state.name)
colnames(statestuff) <- c("State","StateName")
state <- full_join(state,statestuff)
state$StateName <- as.character(state$StateName)
state <- state%>%mutate(StateName=ifelse(is.na(StateName),"D.C.",StateName))
team$State <- as.character(team$State)
team <- team%>%mutate(State=ifelse(State=="District of Columbia","D.C.",State))
state <- state%>%mutate(State1=State,State=StateName)
AllData <- right_join(state,team)
AllData <- AllData%>%filter(!is.na(Happiness))
write.csv(AllData,"AllData.csv")

# make strings numeric
AllData$Capacity <- as.numeric(gsub(",", "", AllData$Capacity) )
AllData$FranchiseValue <- as.numeric(gsub(",", "", AllData$FranchiseValue) )

# making percent of home wins as variable

AllData <- AllData %>% mutate(WinPct = HomeWins/HomeGames)
```

# EDA

```{r}
# Team Level Variables
# Basic EDA
# #
boxplot(AllData$WinPct)
summary(AllData$WinPct)
# # Sport
table(AllData$Type)

# # Capacity
hist(AllData$Capacity)
summary(AllData$Capacity)
ggplot(AllData, aes(x=Type, y = Capacity)) + geom_boxplot() 

# # FranchiseValue
hist(AllData$FranchiseValue)
summary(AllData$FranchiseValue)
ggplot(AllData, aes(x=Type, y = FranchiseValue)) + geom_boxplot() 

# # Attendance
hist(AllData$Attendence)
summary(AllData$Attendence)
ggplot(AllData, aes(x=Type, y = Attendence)) + geom_boxplot() 

# # Year
hist(AllData$Year.opened)
tapply(AllData$Capacity, AllData$Year.opened, mean)
tapply(AllData$Year.opened, AllData$Type,mean) # hockey and basketball have newest stadiums and baseball and football have older, with baseball being oldest
```

# Potential Interactions

Significant Looking interactions Level 1:
Capacity and Attandence, sport and franchise, year and sport, sport and attendance, franchise and attendence, sport and capacity
```{r}
# Level 1 Interactions

# make capacity a factor for data exploration
AllData$CapFactor <- cut(AllData$Capacity, breaks = c(0, 18578, 27000,  62250,100000), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))

# make franchise a factor for data exploration
AllData$FranchiseFactor <- cut(AllData$FranchiseValue, breaks = c(0, 1.000e+09, 1.350e+09,  2.300e+09,4.900e+09), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))

# make attendance a factor for data exploration
AllData$YearFactor <- cut(AllData$Year.opened, breaks = c(0, 1994, 1999,  2004,2019), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))


#cap and year
ggplot(AllData, aes(x=Year.opened, y = Capacity, color = WinPct)) + geom_point()+geom_smooth(method=lm)
ggplot(AllData, aes(x=Year.opened,y = WinPct)) + geom_point() + facet_wrap(~CapFactor)+geom_smooth(method=lm) # doesn't appear to be an interaction

#cap and sport
ggplot(AllData, aes(x=Capacity,y = WinPct)) + geom_point() + facet_wrap(~Type)+geom_smooth(method=lm) # Interaction

#cap and franchise $$
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~CapFactor)+geom_smooth(method=lm) # doesn't appear to be an interaction

#cap and attendance
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~CapFactor) +geom_smooth(method=lm) # Interaction

# year and sport
ggplot(AllData, aes(x=Year.opened,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # potentially...

# year and franchise $
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~YearFactor) +geom_smooth(method=lm) # nope

# year and attandence
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~YearFactor) +geom_smooth(method=lm) # nope

# sport and franchise $
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # interaction

# sport and attendance
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # potentially

# franchise $ and attendance
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~FranchiseFactor) +geom_smooth(method=lm) # potentially

```

Significant Looking Interactions Level 2:
*check all* pop and happy, happy and trump, pop and trump
```{r}
# Level 2 interactions

# pop and happy
ggplot(AllData, aes(x=Population,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # interaction!

# happy and trump
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # interaction!

# pop and trump
# make population a factor for data exploration
AllData$PopFactor <- cut(AllData$Population, breaks = c(0, 6782564, 11694664,  21312211,39776831), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~as.factor(PopFactor)) +geom_smooth(method=lm) # interaction!
```

Significant Looking Interactions Cross Levels:

happiness and capacity, happiness and year, hapiness and sport, population and year, trump and attendance

```{r}
# hap and cap
ggplot(AllData, aes(x=Capacity,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # interaction!

# hap and year
ggplot(AllData, aes(x=Year.opened,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # interaction!

# hap and sport
ggplot(AllData, aes(x=Type,y = WinPct)) + geom_boxplot(aes(color = as.factor(Happiness))) # interaction potential

# hap and fran$
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # not really

# hap and att
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # not really

# pop and cap
ggplot(AllData, aes(x=Population,y = WinPct)) + geom_point() + facet_wrap(~CapFactor) +geom_smooth(method=lm) # not really

# pop and year
ggplot(AllData, aes(x=Population,y = WinPct)) + geom_point() + facet_wrap(~YearFactor) +geom_smooth(method=lm) # potentially

# pop and sport
ggplot(AllData, aes(x=Population,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # not really

# pop and fran$
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~PopFactor) +geom_smooth(method=lm) # not really

# pop and att
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~PopFactor) +geom_smooth(method=lm) # no

# trump and cap
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~CapFactor) +geom_smooth(method=lm) # no

# trump and year
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~YearFactor) +geom_smooth(method=lm) # no

# trump and sport
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # no

# trump and fran$
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~FranchiseFactor) +geom_smooth(method=lm) # no

# trump and att
AllData$TrumpFactor <- cut(AllData$Trump.Vote, breaks = c(0, 36.50,47.25,  50.40,65.40), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~TrumpFactor) +geom_smooth(method=lm) # potentially

```


#Model Building
```{r}
#Scaling Variables
AllData$Happiness2 <- as.factor(AllData$Happiness)
AllData <- AllData%>% mutate(Capacity2=scale(Capacity),Population2=scale(Population),FranchiseValue2=scale(FranchiseValue),Attendence2=scale(Attendence),Trump.Vote2=scale(Trump.Vote),Year.opened2=scale(Year.opened))

#Random Intercept Check
sports.glmerNone <- glmer(WinPct~1+(1|State),family=binomial,weights=HomeGames,data=AllData) #Standard Deviation for random effect is about 0.1909 indicating that there is variation within states related to win percentage. This seems small, but considering win percentage is between 0 and 1, this indicates a standard devation in games won of 19.09%
summary(sports.glmerNone)

#Model will all interactions that looked potentially influential - Full model
sports.glmerFULL <- glmer(WinPct~Capacity2+Attendence2+Type+FranchiseValue2+Year.opened2+Capacity2:Attendence2+Type:FranchiseValue2+Year.opened2:Type+Type:Attendence2+FranchiseValue2:Attendence2+Type:Capacity2+(Population2+Trump.Vote2+Happiness)^2+Happiness:Capacity2+Trump.Vote2:Attendence2+Happiness:Type+Population2:Year.opened2+Happiness:Year.opened2+(1|State),family=binomial,data=AllData,weights=HomeGames)
summary(sports.glmerFULL)

#Starting Build of model with individual terms
sports.glmer <- glmer(WinPct~Capacity2+Attendence2+Type+FranchiseValue2+Year.opened2+Population2+Trump.Vote2+Happiness+(1|State),family = binomial,data=AllData,weights=HomeGames)
summary(sports.glmer)

#From full model looked like type:attendence, type:capacity, population:happiness, population:trump, and happiness:year looked most influential
sports.glmer2 <- update(sports.glmer,.~.+Type:Attendence2+Type:Capacity2+Population2:Happiness+Population2:Trump.Vote2+Happiness:Year.opened2)
anova(sports.glmer,sports.glmer2) #Very low p value - at least one of the 5 additional terms is significant

summary(sports.glmer2) #population2:Trump and year:happiness look non significant

sports.glmer3 <- update(sports.glmer2,.~.-Population2:Trump.Vote2-Year.opened2:Happiness)
anova(sports.glmer2,sports.glmer3) #neither significant in the model
summary(sports.glmer3) #looks like trump.vote and year.opened not significant

sports.glmer4 <- update(sports.glmer3,.~.-Trump.Vote2-Year.opened2)
anova(sports.glmer4,sports.glmer3) #smaller prefered
summary(sports.glmer4)

#Now we will try to add in other significant interactions from EDA

#add capacity:attendence
sports.glmer5 <- update(sports.glmer4,.~.+Capacity2:Attendence2)
anova(sports.glmer4,sports.glmer5) #larger prefered
summary(sports.glmer5)


#add type:franchise.value
sports.glmer6 <- update(sports.glmer5,.~.+Type:FranchiseValue2)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)


#add year:type add in year to test
sports.glmer6 <- update(sports.glmer5,.~.+Year.opened2+Year.opened2:Type)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#franchise:attendence
sports.glmer6 <- update(sports.glmer5,.~.+Attendence2:FranchiseValue2)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#Happiness:trump
sports.glmer6 <- update(sports.glmer5,.~.+Happiness:Trump.Vote2)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#Happiness:capacity
sports.glmer6 <- update(sports.glmer5,.~.+Happiness:Capacity2)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#Happiness:Type
sports.glmer6 <- update(sports.glmer5,.~.+Type:Happiness)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#Population:Year - add year back in
sports.glmer6 <- update(sports.glmer5,.~.+Year.opened2+Year.opened2:Population2)
anova(sports.glmer6,sports.glmer5) #larger prefered
summary(sports.glmer6) #interaction actually not significant

#remove population:year
sports.glmer7 <- update(sports.glmer6,.~.-Year.opened2:Population2)
anova(sports.glmer6,sports.glmer7) #smaller prefered
summary(sports.glmer7)

#trump:attendence
sports.glmer8 <- update(sports.glmer7,.~.+Trump.Vote2+Trump.Vote2:Attendence2)
anova(sports.glmer8,sports.glmer7) #smaller prefered
summary(sports.glmer8)

# franchise value is marginally significant, let's try to take it out
#trump:attendence
sports.glmer8 <- update(sports.glmer7,.~.-FranchiseValue2)
anova(sports.glmer8,sports.glmer7) #it's pvalue is so close to .05 let's leave it
summary(sports.glmer8)

#Final Model - Model 7
sports.glmerFinal <- sports.glmer7
summary(sports.glmerFinal)
```

Here are the effects of each variable
```{r}
# Capacity interpretations
sd(AllData$Capacity)

# increasing capacity by 23,200 people increases the odds of winning at home by 3.36 times for baseball. However this number doesn't make sense since we have such a range in values. Therefore we need to do it by a fraction 

#baseball SE is 1039.8813 so 1039.8813/23200=0.0448
exp(1.21418*.263) # increase baseball stadium by 6,116 people 37% increase
#basketball SE is 6116.0482 so 6116.0482/23,200=.263
exp(1.21418*0.04482247 +  0.97581) # basketball increasing capacity increases the odds of winning by 2.8 times
#football SE is 11875.7172 so 11875.7172/23200=.511
exp(1.21418*0.5118844 +  -1.85881) # football increasing capacity decreases the odds of winning by 70.9%
#hockey SE is 998.3133 so 998.3133/23200=0.043
exp(1.21418*0.043 +  -5.12141) # hockey increasing capacity decreases the odds of winning by 99%


# Type Sport Interpretations
exp(2.38943) # basketball has 10.9 times more likely to win in home court than baseball
exp(1.09150) # football is 3 times more likely to win in home court than baseball
exp(-3.19716) # hockey is 96% less likely to win in home court than baseball

# makes sense because baseball stadiums are all different shapes, maybe hockey you cannot hear fans so home court doesn't matter

# Attendence Interpretations
sd(AllData$Attendence)

# increasing attendance by 18.12 percent decreases the odds of winning at home by 6% but this changes based on sport and capacity. So if Baseball is the baseline, 6%
exp(-0.06179)
exp(-0.06179+ 0.32640) # basketball 30% increase
exp(-0.06179+-0.85513) # football 60% decrease
exp(-0.06179+0.88390) # hockey is 2.3 times increase

# interaction between capacity and attendance
exp(0.62979*.263) # 1.2 times more baseball, so increase one and the effect of the other increases by 20%
exp(0.62979*0.04482247) # 1.02 times more basketball, so increase one and the effect of the other increases by 2%
exp(0.62979*0.5118844) # 1.38 times more football, so increase one and the effect of the other increases by 38%
exp(0.62979*0.043) # 1.027 times more hockey, so increase one and the effect of the other increases by 2.7%


# Franchise value interpretations
sd(AllData$FranchiseValue)
# increasing franchise value by 93,5646,568 increases odds of winning at home by 14%
exp(0.13274) 

# population interpretations
sd(AllData$Population)
# increasing a state's population by 11,921,447 decreases the odds of a home win by 62% but don't forget about interaction
exp(-0.96189)

# interaction is 24% increase, so increases happiness effect by 24%
exp(0.21383)

# happy interpretation
# increasing happiness by 1, increases odds of winning at home by 17%
exp(0.15703)

# year opened
# increasing the year a stadium opened decreases the odds of winning by 8%, potentially older well established teams have better winning records while newer teams with newer stadiums do not have same legacy
exp(-0.08524)
```


Using the model results, the most significant predictors of home winning percentag were stadium capacity, state population, state happiness level, attendence for hockey games, and type of sport. Many of these variables relate to characteristics of the stadium as well as a team's fan base. The only variable that was completely removed from the model was percent of Trump vote meaning that the percent of people in a state who voted for Trump had no effect on the home winning percentage of a team. 

First we will examine how team level factors effect home win percentage. The first is what sport we want to look at. Compared to baseball teams (our baseline) basketball teams have 10.9 times higher odds of winning at home and football is 3 times more likely to win at home than baseball. Hockey on the other hand sees a decrease in home winning odds of 96% compared to baseball. 

For baseball teams, increasing their stadium capacity by 1,039 (the standard deviation of stadium capacity for baseball teams) increased the mean odds of winning at home by 37%. For basketball teams the effect is much stronger. Increasing capacity by their capacity standard error (6,116 seats) increases mean odds of winning at home by 2.8 times. Football teams actually see a decrease of 70.9% in mean home winning odds when they increase their stadium capacity by a standard deviation (11875 seats). Finally, hockey sees a decrease as well with their home game winning odds decreasing by 99% when they increase their stadium capacity one standard error of 998 seats. Increasing stadium size also chances the effect of attendence on odds of home wins since there is an interaction term for these variables. Increasing the baseball stadium capacity by the above amount increases the effect of attendance by 20%. Increasing basketball capacity increases the effect of attendance by 2%, football capacity increases attendence's effect by 38% and hockey 2.7%. Increasing a team's franchise value by $93,5646,568 (the standard deviation of this data) increases odds of winning at home by 14% holding all else constant. Finally, newer stadiums actually decrease the odds of winning. For every 19.8 years newer a stadium is, the mean odds of winning decreases by 8% holding all else constant. *for discussion potentially older well established teams have better winning records while newer teams with newer stadiums do not have same legacy * Attendence's effect varies by type of sport. Increasing attendence by 18.12 percentage points, decreases the mean odds of home wins by 6% for baseball teams. Increasing attendence by 18.12 percentage points also decreases the odds of home winning for footbal but by 60% holding all else constant. For basketball teams, this chance increases mean odds by 60% and for hockey, teams odds of winning are 2.3 times higher. *for disscussion this could be because these are smaller more intimate venues so attendence matters more*

Moving onto state level effects, increasing a state's population by one standard deviation (11,921,447 people) decreases the odds of home wins by 62% but also has an implication on the effect of happiness since there is an interaction term. Increasing happiness level by 1, increases the mean odds of winning at home by 17% holding all else constant. We can interpret the interaction between happiness and population by if we increase populaton by 11,921,447 people, the effect of happiness increases by 24%. 



