---
title: "Code"
author: "Clara Livingston & Emily Kaegi"
date: "May 25, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(lme4)
library(fiftystater)
library(stargazer)
library(gridExtra)
```

#Data Join
```{r}
state <- read.csv("StateData.csv")
team <- read.csv("TeamData2.csv",stringsAsFactors = TRUE)

statestuff <- data.frame(state.abb,state.name)
colnames(statestuff) <- c("State","StateName")
state <- full_join(state,statestuff)
state$StateName <- as.character(state$StateName)
state <- state%>%mutate(StateName=ifelse(is.na(StateName),"D.C.",StateName))
team$State <- as.character(team$State)
team <- team%>%mutate(State=ifelse(State=="District of Columbia","D.C.",State))
state <- state%>%mutate(State1=State,State=StateName)
AllData <- right_join(state,team)
AllData <- AllData%>%filter(!is.na(Happiness))

# make strings numeric
AllData$Capacity <- as.numeric(gsub(",", "", AllData$Capacity) )
AllData$FranchiseValue <- as.numeric(gsub(",", "", AllData$FranchiseValue) )

AllData$Happiness2 <- as.factor(AllData$Happiness)

AllData <- AllData%>% mutate(Capacity2=scale(Capacity),Population2=scale(Population),FranchiseValue2=scale(FranchiseValue),Attendence2=scale(Attendence),Trump.Vote2=scale(Trump.Vote),Year.opened2=scale(Year.opened))

# making percent of home wins as variable

AllData <- AllData %>% mutate(WinPct = HomeWins/HomeGames)
write.csv(AllData,"AllData.csv")
```

# EDA

```{r}
# Team Level Variables
# Basic EDA
# #
boxplot(AllData$WinPct)
summary(AllData$WinPct)
# # Sport
table(AllData$Type)

# # Capacity
hist(AllData$Capacity)
summary(AllData$Capacity)
ggplot(AllData, aes(x=Type, y = Capacity)) + geom_boxplot() 

# # FranchiseValue
hist(AllData$FranchiseValue)
summary(AllData$FranchiseValue)
ggplot(AllData, aes(x=Type, y = FranchiseValue)) + geom_boxplot() 

# # Attendance
hist(AllData$Attendence)
summary(AllData$Attendence)
ggplot(AllData, aes(x=Type, y = Attendence)) + geom_boxplot() 

# # Year
hist(AllData$Year.opened)
tapply(AllData$Capacity, AllData$Year.opened, mean)
tapply(AllData$Year.opened, AllData$Type,mean) # hockey and basketball have newest stadiums and baseball and football have older, with baseball being oldest


```

EDA for State Level Variables
```{r,fig.cap="Caption Goes Here",results='asis',warning=FALSE}
# states with teams
length(unique(AllData$State))

#Variation between states win pct
AllStates <- fifty_states
AllStates$state <- AllStates$id
state <- read.csv("StateData.csv")
statestuff <- data.frame(state.abb,state.name)
colnames(statestuff) <- c("State","StateName")
state <- full_join(state,statestuff)
state$StateName <- as.character(state$StateName)
state <- state%>%mutate(StateAbb=State,State=ifelse(State=="DC","District of Columbia",StateName))
state <- state%>%mutate(state2=tolower(State))
AllData <- read.csv("AllData.csv")
AllData <- AllData%>%mutate(state2=tolower(State),WinPct = HomeWins/HomeGames)
AllData2 <- AllData%>%group_by(state2)%>%summarise(AvgWin=mean(WinPct))

#Population
ggplot() + coord_map() + geom_map(data = state, aes(map_id = state2, fill = Population), map = AllStates) + expand_limits(x = AllStates$long, y = AllStates$lat) + ggtitle("Population By State 2017") + labs(x= "Longitute", y  = "Latitude") +scale_fill_distiller(name = "Population", palette = "BuPu") +theme_classic()

#Happiness
ggplot() + coord_map() + geom_map(data = state, aes(map_id = state2, fill = Happiness), map = AllStates) + expand_limits(x = AllStates$long, y = AllStates$lat) + ggtitle("Happiness By State 2017") + labs(x= "Longitute", y  = "Latitude") +scale_fill_distiller(name = "Happiness", palette = "BuPu") +theme_classic()

#Trump Vote
ggplot() + coord_map() + geom_map(data = state, aes(map_id = state2, fill = Trump.Vote), map = AllStates) + expand_limits(x = AllStates$long, y = AllStates$lat) + ggtitle("Trump Vote By State 2017") + labs(x= "Longitute", y  = "Latitude") +scale_fill_distiller(name = "Vote", palette = "BuPu") +theme_classic()
```

# Potential Interactions

Significant Looking interactions Level 1:
Capacity and Attandence, sport and franchise, year and sport, sport and attendance, franchise and attendence, sport and capacity
```{r}
# Level 1 Interactions

# make capacity a factor for data exploration
AllData$CapFactor <- cut(AllData$Capacity, breaks = c(0, 18578, 27000,  62250,100000), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))

# make franchise a factor for data exploration
AllData$FranchiseFactor <- cut(AllData$FranchiseValue, breaks = c(0, 1.000e+09, 1.350e+09,  2.300e+09,4.900e+09), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))

# make attendance a factor for data exploration
AllData$YearFactor <- cut(AllData$Year.opened, breaks = c(0, 1994, 1999,  2004,2019), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))


#cap and year
ggplot(AllData, aes(x=Year.opened, y = Capacity, color = WinPct)) + geom_point()+geom_smooth(method=lm)
ggplot(AllData, aes(x=Year.opened,y = WinPct)) + geom_point() + facet_wrap(~CapFactor)+geom_smooth(method=lm) # doesn't appear to be an interaction

#cap and sport
ggplot(AllData, aes(x=Capacity,y = WinPct)) + geom_point() + facet_wrap(~Type)+geom_smooth(method=lm) # Interaction

#cap and franchise $$
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~CapFactor)+geom_smooth(method=lm) # doesn't appear to be an interaction

#cap and attendance
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~CapFactor) +geom_smooth(method=lm) # Interaction

# year and sport
ggplot(AllData, aes(x=Year.opened,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # potentially...

# year and franchise $
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~YearFactor) +geom_smooth(method=lm) # nope

# year and attandence
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~YearFactor) +geom_smooth(method=lm) # nope

# sport and franchise $
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # interaction

# sport and attendance
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # potentially

# franchise $ and attendance
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~FranchiseFactor) +geom_smooth(method=lm) # potentially

```

Significant Looking Interactions Level 2:
*check all* pop and happy, happy and trump, pop and trump
```{r}
# Level 2 interactions

# pop and happy
ggplot(AllData, aes(x=Population,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # interaction!

# happy and trump
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # interaction!

# pop and trump
# make population a factor for data exploration
AllData$PopFactor <- cut(AllData$Population, breaks = c(0, 6782564, 11694664,  21312211,39776831), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~as.factor(PopFactor)) +geom_smooth(method=lm) # interaction!
```

Significant Looking Interactions Cross Levels:

happiness and capacity, happiness and year, hapiness and sport, population and year, trump and attendance

```{r}
# hap and cap
ggplot(AllData, aes(x=Capacity,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # interaction!

# hap and year
ggplot(AllData, aes(x=Year.opened,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # interaction!

# hap and sport
ggplot(AllData, aes(x=Type,y = WinPct)) + geom_boxplot(aes(color = as.factor(Happiness))) # interaction potential

# hap and fran$
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # not really

# hap and att
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~as.factor(Happiness)) +geom_smooth(method=lm) # not really

# pop and cap
ggplot(AllData, aes(x=Population,y = WinPct)) + geom_point() + facet_wrap(~CapFactor) +geom_smooth(method=lm) # not really

# pop and year
ggplot(AllData, aes(x=Population,y = WinPct)) + geom_point() + facet_wrap(~YearFactor) +geom_smooth(method=lm) # potentially

# pop and sport
ggplot(AllData, aes(x=Population,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # not really

# pop and fran$
ggplot(AllData, aes(x=FranchiseValue,y = WinPct)) + geom_point() + facet_wrap(~PopFactor) +geom_smooth(method=lm) # not really

# pop and att
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~PopFactor) +geom_smooth(method=lm) # no

# trump and cap
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~CapFactor) +geom_smooth(method=lm) # no

# trump and year
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~YearFactor) +geom_smooth(method=lm) # no

# trump and sport
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~Type) +geom_smooth(method=lm) # no

# trump and fran$
ggplot(AllData, aes(x=Trump.Vote,y = WinPct)) + geom_point() + facet_wrap(~FranchiseFactor) +geom_smooth(method=lm) # no

# trump and att
AllData$TrumpFactor <- cut(AllData$Trump.Vote, breaks = c(0, 36.50,47.25,  50.40,65.40), labels = c("1st Q", "2nd Q", "3rd Q", "4th Q"))
ggplot(AllData, aes(x=Attendence,y = WinPct)) + geom_point() + facet_wrap(~TrumpFactor) +geom_smooth(method=lm) # potentially

```


#Model Building
```{r}
#Scaling Variables
AllData$Happiness2 <- as.factor(AllData$Happiness)

AllData <- AllData%>% mutate(Capacity2=scale(Capacity),Population2=scale(Population),FranchiseValue2=scale(FranchiseValue),Attendence2=scale(Attendence),Trump.Vote2=scale(Trump.Vote),Year.opened2=scale(Year.opened))

#Random Intercept Check
sports.glmerNone <- glmer(WinPct~1+(1|State),family=binomial,weights=HomeGames,data=AllData) #Standard Deviation for random effect is about 0.1909 indicating that there is variation within states related to win percentage. This seems small, but considering win percentage is between 0 and 1, this indicates a standard devation in games won of 19.09%
summary(sports.glmerNone)

#Model will all interactions that looked potentially influential - Full model
sports.glmerFULL <- glmer(WinPct~Capacity2+Attendence2+Type+FranchiseValue2+Year.opened2+Capacity2:Attendence2+Type:FranchiseValue2+Year.opened2:Type+Type:Attendence2+FranchiseValue2:Attendence2+Type:Capacity2+(Population2+Trump.Vote2+Happiness)^2+Happiness:Capacity2+Trump.Vote2:Attendence2+Happiness:Type+Population2:Year.opened2+Happiness:Year.opened2+(1|State),family=binomial,data=AllData,weights=HomeGames)
summary(sports.glmerFULL)

#Starting Build of model with individual terms
sports.glmer <- glmer(WinPct~Capacity2+Attendence2+Type+FranchiseValue2+Year.opened2+Population2+Trump.Vote2+Happiness+(1|State),family = binomial,data=AllData,weights=HomeGames)
summary(sports.glmer)

#From full model looked like type:attendence, type:capacity, population:happiness, population:trump, and happiness:year looked most influential
sports.glmer2 <- update(sports.glmer,.~.+Type:Attendence2+Type:Capacity2+Population2:Happiness+Population2:Trump.Vote2+Happiness:Year.opened2)
anova(sports.glmer,sports.glmer2) #Very low p value - at least one of the 5 additional terms is significant

summary(sports.glmer2) #population2:Trump and year:happiness look non significant

sports.glmer3 <- update(sports.glmer2,.~.-Population2:Trump.Vote2-Year.opened2:Happiness)
anova(sports.glmer2,sports.glmer3) #neither significant in the model
summary(sports.glmer3) #looks like trump.vote and year.opened not significant

sports.glmer4 <- update(sports.glmer3,.~.-Trump.Vote2-Year.opened2)
anova(sports.glmer4,sports.glmer3) #smaller prefered
summary(sports.glmer4)

#Now we will try to add in other significant interactions from EDA

#add capacity:attendence
sports.glmer5 <- update(sports.glmer4,.~.+Capacity2:Attendence2)
anova(sports.glmer4,sports.glmer5) #larger prefered
summary(sports.glmer5)


#add type:franchise.value
sports.glmer6 <- update(sports.glmer5,.~.+Type:FranchiseValue2)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)


#add year:type add in year to test
sports.glmer6 <- update(sports.glmer5,.~.+Year.opened2+Year.opened2:Type)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#franchise:attendence
sports.glmer6 <- update(sports.glmer5,.~.+Attendence2:FranchiseValue2)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#Happiness:trump
sports.glmer6 <- update(sports.glmer5,.~.+Happiness:Trump.Vote2)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#Happiness:capacity
sports.glmer6 <- update(sports.glmer5,.~.+Happiness:Capacity2)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#Happiness:Type
sports.glmer6 <- update(sports.glmer5,.~.+Type:Happiness)
anova(sports.glmer6,sports.glmer5) #smaller prefered
summary(sports.glmer6)

#Population:Year - add year back in
sports.glmer6 <- update(sports.glmer5,.~.+Year.opened2+Year.opened2:Population2)
anova(sports.glmer6,sports.glmer5) #larger prefered
summary(sports.glmer6) #interaction actually not significant

#remove population:year
sports.glmer7 <- update(sports.glmer6,.~.-Year.opened2:Population2)
anova(sports.glmer6,sports.glmer7) #smaller prefered
summary(sports.glmer7)

#trump:attendence
sports.glmer8 <- update(sports.glmer7,.~.+Trump.Vote2+Trump.Vote2:Attendence2)
anova(sports.glmer8,sports.glmer7) #smaller prefered
summary(sports.glmer8)

# franchise value is marginally significant, let's try to take it out
#trump:attendence
sports.glmer8 <- update(sports.glmer7,.~.-FranchiseValue2)
anova(sports.glmer8,sports.glmer7) #it's pvalue is so close to .05 let's leave it
summary(sports.glmer8)

#Final Model - Model 7
sports.glmerFinal <- sports.glmer7
summary(sports.glmerFinal)
saveRDS(sports.glmerFinal, "Final.rds")
```

Here are the effects of each variable
```{r}
# Capacity interpretations
sd(AllData$Capacity)

# increasing capacity by 23,200 people increases the odds of winning at home by 3.36 times for baseball. However this number doesn't make sense since we have such a range in values. Therefore we need to do it by a fraction 

#baseball SE is 1039.8813 so 1039.8813/23200=0.0448
exp(1.21418*.263) # increase baseball stadium by 6,116 people 37% increase
#basketball SE is 6116.0482 so 6116.0482/23,200=.263
exp(1.21418*0.04482247 +  0.97581) # basketball increasing capacity increases the odds of winning by 2.8 times
#football SE is 11875.7172 so 11875.7172/23200=.511
exp(1.21418*0.5118844 +  -1.85881) # football increasing capacity decreases the odds of winning by 70.9%
#hockey SE is 998.3133 so 998.3133/23200=0.043
exp(1.21418*0.043 +  -5.12141) # hockey increasing capacity decreases the odds of winning by 99%


# Type Sport Interpretations
exp(2.38943) # basketball has 10.9 times more likely to win in home court than baseball
exp(1.09150) # football is 3 times more likely to win in home court than baseball
exp(-3.19716) # hockey is 96% less likely to win in home court than baseball

# makes sense because baseball stadiums are all different shapes, maybe hockey you cannot hear fans so home court doesn't matter

# Attendence Interpretations
sd(AllData$Attendence)

# increasing attendance by 18.12 percent decreases the odds of winning at home by 6% but this changes based on sport and capacity. So if Baseball is the baseline, 6%
exp(-0.06179)
exp(-0.06179+ 0.32640) # basketball 30% increase
exp(-0.06179+-0.85513) # football 60% decrease
exp(-0.06179+0.88390) # hockey is 2.3 times increase

# interaction between capacity and attendance
exp(0.62979*.263) # 1.2 times more baseball, so increase one and the effect of the other increases by 20%
exp(0.62979*0.04482247) # 1.02 times more basketball, so increase one and the effect of the other increases by 2%
exp(0.62979*0.5118844) # 1.38 times more football, so increase one and the effect of the other increases by 38%
exp(0.62979*0.043) # 1.027 times more hockey, so increase one and the effect of the other increases by 2.7%


# Franchise value interpretations
sd(AllData$FranchiseValue)
# increasing franchise value by 93,5646,568 increases odds of winning at home by 14%
exp(0.13274) 

# population interpretations
sd(AllData$Population)
# increasing a state's population by 11,921,447 decreases the odds of a home win by 62% but don't forget about interaction
exp(-0.96189)

# interaction is 24% increase, so increases happiness effect by 24%
exp(0.21383)

# happy interpretation
# increasing happiness by 1, increases odds of winning at home by 17%
exp(0.15703)

# year opened
# increasing the year a stadium opened decreases the odds of winning by 8%, potentially older well established teams have better winning records while newer teams with newer stadiums do not have same legacy
exp(-0.08524)
```






