---
title: "Influence of Stadium and State Characteristics on Home Winning Percentage"
author: "Clara Livingston and Emily Kaegi"
date: "May 25, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fiftystater)
library(ggplot2)
```

#Introduction

Typically, when looking at statistics related to sports, researchers tend to focus on the dynamics of the team in question when predicting winning percentages. The goal of our research was to take analysis away from factors related to the team and investigate how factors relating to the location of the team's home stadium affects their home winning percentage. Most people are aware of the phrase "homefield advantage", but just how influential is a team's homefield in boosting their win percentages? SBNation and Sporting News both claim that home field advantage is real across most sports, especially the NBA. What appears harder to determine, however, is the cause of variability in this percentage between different sports and teams. 

#Methods

#Results

#Discussion

#Appendix




```{r,echo=FALSE}
AllStates <- fifty_states
AllStates$state <- AllStates$id
state <- read.csv("StateData.csv")
statestuff <- data.frame(state.abb,state.name)
colnames(statestuff) <- c("State","StateName")
state <- full_join(state,statestuff)
state$StateName <- as.character(state$StateName)
state <- state%>%mutate(StateAbb=State,State=ifelse(State=="DC","District of Columbia",StateName))
state <- state%>%mutate(state2=tolower(State))
AllData <- read.csv("AllData.csv")
AllData <- AllData%>%mutate(state2=tolower(State),WinPct = HomeWins/HomeGames)
AllData2 <- AllData%>%group_by(state2)%>%summarise(AvgWin=mean(WinPct))

#Population
ggplot() + coord_map() + geom_map(data = state, aes(map_id = state2, fill = Population), map = AllStates) + expand_limits(x = AllStates$long, y = AllStates$lat) + ggtitle("Population By State 2017") + labs(x= "Longitute", y  = "Latitude") +scale_fill_distiller(name = "Population", palette = "BuPu") +theme_classic()

#Happiness
ggplot() + coord_map() + geom_map(data = state, aes(map_id = state2, fill = Happiness), map = AllStates) + expand_limits(x = AllStates$long, y = AllStates$lat) + ggtitle("Happiness By State 2017") + labs(x= "Longitute", y  = "Latitude") +scale_fill_distiller(name = "Happiness", palette = "BuPu") +theme_classic()

state2 <- state%>%filter(StateAbb!="DC")
#Average percent home games won by state
ggplot() + coord_map() + geom_map(data = AllData2, aes(map_id = state2, fill = AvgWin), map = AllStates) + expand_limits(x = AllStates$long, y = AllStates$lat) + ggtitle("Percent Won By State") + labs(x= "Longitute", y  = "Latitude") +scale_fill_distiller(name = "Percent", palette = "BuPu") +theme_classic()
```
